---
title: "Doing Data Science Case Study 1 Unit 6, Week 6"
author: "Najeeb Zaidi"
date: "July 27, 2016"
output: html_document
---

# World Bank Country Data : Comparative Analysis of GDP vs ....

## Introduction 

The analysis attempts to profile 190 Countries based on their respective GDP and explore :


1	Match the data based on the country shortcode. How many of the IDs match? 

2	Sort the data frame in ascending order by GDP rank (so United States is last). What is the 13th country in the resulting data frame?

3	What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups? 

4	Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.

5	Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?


The data sources used for the presentation are:

    * http://data.worldbank.org/data-catalog/GDP-ranking-table
    * http://data.worldbank.org/data-catalog/ed-stats

## Library Tools

We have used the RMarkdown and RStudio to create this HTML and to ensure version control and standards of reproducible research. The R package libraries used for this purpose are:

     * library(mosaic)
     * library(stats)
     * library(ggplot2)
     * library(data.table)

## Getting Data from identified sources

We first set our desired work directory for saving the workfile and check it by repeating the command 

```{r include=FALSE, echo=FALSE}

library(ggplot2)
library(stats)
library(mosaic)
library(data.table)

rm(list = ls()) 
objects()
getwd() 
setwd("C:\\Users\\najee\\Documents\\R\\MSDS6306-W6-HW")
```

Data Source 1: procure and read data from target source
```{r echo = FALSE}
GDPData <- read.csv("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv ", 
                    header=T, sep=",", 
                    stringsAsFactors = FALSE)
GDPData2 <- GDPData[-c(1:4,195:330), -c(3,6,7:10)]
names(GDPData2) <- c("ID","Ranking","Country", "GDP")
dim(GDPData2)
str(GDPData2)
```

Data Source 2: procure and read data from target source
```{r }
EduData <- read.csv("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv", 
                    header = TRUE, sep = ",", 
                    stringsAsFactors = FALSE)
names(EduData) <- c("ID","Country","Income_Group", "Region", "Lending_Group", "Other")
dim(EduData)
```
Question 1: Match the data based on the country shortcode. How many of the IDs match?  
Merging and removing duplicate country names from the Data merged dataset

```{r}
Data <- merge(GDPData2, EduData, by = "ID") 
attach(Data)
dim(Data)
answ1 <- length(GDP)
answ1
message("Number of countries for which both data type are available is: 189")
```

Question 2: Sort the data frame in ascending order by GDP rank. What is the 13th country in the resulting data frame?  
Ensuring GDP is captured as numeric vector and is not in scientific format in the dataframe (for whatever reason this step might be implemented twice to have effect)

```{r}
sapply(Data,class) 
typeof(GDP)
Data$GDP <- gsub("\\,","",GDP)
Data2 <- transform(Data, GDP = as.numeric(as.character(GDP)))
typeof(GDP)
# Sorting data frame as requested and displaying and storing answer 2
sort.Data2 <- Data2[order(Data2$GDP), ]
attach(sort.Data2)
sort.Data2[13,3] # Gives the names of the 13th country in the merged list by GDP asc.
answ2 <- sort.Data2[13,3] 
message("The 13th country by GDP (ascending) is: St. Kitts and Nevis")
```
Question 3: What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?  
Ensuring also the Ranking attribute is captured as numeric vector and verifying that no further missing values are for the attributes in scope for the rest of the exercise

```{r}
sapply(sort.Data2,class) 
Data3 <- transform(sort.Data2, Ranking = as.numeric(as.character(Ranking)))
sapply(Data3,class)
attach(Data3)

#Displaying and storing answer(s) 3
answ3.1 <- with(Data3, mean(Ranking[Income_Group == "High income: OECD"]))
answ3.1
message("The avg Ranking for the ?High income: OECD? countries is: 32.9")

answ3.2 <- with(Data3, mean(Ranking[Income_Group == "High income: OECD"]))
answ3.2
message("The avg Ranking for the ?High income: nonOECD? countries is: 91.9")

```
Question 4: Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group

```{r}
plot(GDP, bty = "l", col = "green", pch = 19, 
     main = "GDP distribution and avrgs.", 
     xlab = "Countries (asc. Order)", 
     ylab = "GDP in $m")

# Visualizing Income Groups by color descriptions
colour <- c("grey", "yellow","orange","red","black")
col.list <- rep(0, length(Data3$Income_Group))
col.list[Data3$Income_Group == "Low income"] <- 1
col.list[Data3$Income_Group == "Lower middle income"] <- 2
col.list[Data3$Income_Group == "Upper middle income"] <- 3
col.list[Data3$Income_Group == "High income: nonOECD"] <- 4
col.list[Data3$Income_Group == "High income: OECD"] <- 5

p <- plot(GDP, pch = 19, main = "GDP per Income Group", 
           xlab = "Countries (asc. Order)", 
           ylab = "GDP in $m", 
           col = c(colour[col.list]))

legend(x = "topleft", cex = 0.7, c(" Low income in Grey ",
                               " Lower middle income in Yellow ",
                               " Upper middle income in Orange ",
                               " High income ? non OECD in Red ",
                               " High income ? OECD in Black "))

```
Question 5: Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?
Selecting only relevant attributes

```{r}
Data4 <- Data3[c(2,6)]
sort.Data4 <- Data4[order(Ranking), ]

# Computing Ranking quintile and grouping by Income_Group
z = quantile(Ranking,c(0.2,0.4,0.6,0.8),na.rm = FALSE,names = FALSE)

z
# Checking calculated quantiles

b <- setDT(sort.Data4)[Ranking <= z[1], quantile := "1st"][Ranking >= z[1] & Ranking <= z[2], quantile := "2nd"][Ranking >= z[2] & Ranking<=z[3], quantile:= "3rd"][Ranking >= z[3] & Ranking<=z[4], quantile:= "4th"][Ranking >= z[4], quantile:= "5th"]
sapply(b,class) 
b1 = subset(b, quantile == "1st")
b2 = subset(b1, Income_Group == "Lower middle income")

dim(b2) #  This gives the number of Lower middle income countries in the first Ranking quantile (eg 5)
# Display and storing answer to last question 5
answ5 <- dim(b2) 
message("Number of Lower middle income countries in 1st Ranking quantile is: 5")

```
## Summary & Conclusion

We have been able to analyze these countries .....


### Annexes

A. List of countries excluded from original dataset
B. Data cleansing items
C. General description of data trasformation process

